### 2010-10-22 Thesis/Chapter2.tex
- Update 22 October -
Almost done...



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@188 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\section{Description of motion} 	\label{chap2:descriptionMotion}
Let us consider a body $\textbf{\textit{B}}$ of known geometry in a three-dimensional Euclidian space $\field{R}^3$. For a given geometry and loading, $\textbf{\textit{B}}$ will undergo a set of macroscopic changes which is called \emph{deformation}. The region of space occupied by the body at a given time $t$ is termed a \emph{configuration}. A change in the configuration of a continuum body results in a displacement. The displacement of a body has two components: a rigid-body displacement and a deformation. A rigid-body displacement consists of a simultaneous translation and rotation of the body without changing its shape or size. In constrast, a change in shape and\slash or size of the body from an initial configuration $\kappa_{0}$ to a new configuration $\kappa$ is called a deformation. This new configuration $\kappa$ may be designated by the \emph{current} or \emph{deformed configuration}. \OFF

@@

	\subsection{Deformation gradient tensor}
The displacement field tells us how a particule displaces from the reference to the deformed configuration. More interestingly, we would like to know how a material line would deform (stretch and rotation) within this displacement field. Since the length of a material line $d\mathbf{X}$ can change when going to the deformed configuration as well as its orientation, we can say that $d\mathbf{X}$ deforms into $d\mathbf{x}$. We now seek the relation existing between $d\mathbf{x}$ in the deformed configuration and $d\mathbf{X}$ of the reference configuration. \OFF

@@

\section{Constitutive equations}
So far, all the relations that we have established are valid for every continuum. Indeed, the derivations carried out so far made no mention of any material. Therefore, we know that these relations are not sufficient to describe the response of a material to a given loading. Moreover, because the stresses result from the deformation of the material, they may be expressed in terms of some measure of this deformation such as the strain. This relationship, which obviously depends on the type of material under consideration, is what we call a \emph{constitutive equation}. Constitutives equations are an attempt to mathematically characterise the behaviour of materials observed from experimentals results. \OFF

@@

In addition, if there exists a \emph{strain energy density} function $U_0(\varepsilon)$ such that 
\begin{equation}
\label{chap2:energy}
\boldsymbol \sigma = \pd{U_0}{\varepsilon},
\end{equation}
the material is said to be hyperelastic. However, for \emph{imcompressible} materials the stress tensor is not completely determined by the deformation. The hydrostatic pressure\footnote{pressure exerted by a fluid at equilibrium due to the force of gravity} affects the stress and the above equation is then written as:
\begin{equation}
\label{chap2:energy2}
\boldsymbol \sigma = -p \mathbf{I} + \pd{U_0}{\varepsilon},
\end{equation}
where $p$ is the hydrostatic pressure. In order to develop a mathematical model of an hyperelastic material, $U_0$ is expanded in Taylor's series about $\varepsilon = 0$:
\begin{equation}
U_0 = C_0 + C_{ij}\varepsilon_{ij} + \frac{1}{2!}\hat{C}_{ijkl}\varepsilon{ij}\varepsilon_{kl} + \frac{1}{3!}\hat{C}_{ijklmn}\varepsilon_{ij}\varepsilon_{kl}\varepsilon_{mn} + \ldots , 
\end{equation}
where $C_0$, $C_{ij}$ and $\hat{C}$ are material stiffnesses. The form of $U_0$ may varied for different materials. A material is said to be \emph{linear} if the relationship between strain and stress is also linear (see \fig{chap2:fig-linear}). Based on this definition, $U_0$ is a cubic or higher-order function of the strains for non-linear materials whereas $U_0$ is a quadratic function of strain for linear materials. We will start by discussing linear materials for the case of infinitesimal deformations (sections \ref{chap2:Hooke} to \ref{chap2:isotropic}). Consequently, we will not make any distinction between the various measures of stress and strain and use $\boldsymbol \sigma$ and $\boldsymbol \varepsilon$ for stress and strain tensors respectively. 
%
\begin{figure}[ht]
\centering
\includegraphics[width=7cm]{chapter2/linear.pdf}
\caption {The relationship between strain and stress is often non-linear (blue). However, it may be approximated as a linear relationship for small deformations (red). }
\label{chap2:fig-linear}
\end{figure}

@@

Most often, the material properties are determined in a laboratory in terms of the engineering constants such as Young's modulus, shear modulus, and so on. These constants are measured using simple tests like uniaxial tension test or pure shear test. Because of their direct and obvious physical meaning, engineering constants are used in place of the more abstract stiffness coefficients $C_{ijkl}$. Therefore, we will now reformulate \eqref{chap2:orthotropic} with respect to those engineering constants. Noting $E_i$ the Young's modulus in the direction $x_i$ and $G_{ij}$ the shear moduli in the $x_i-x_j$ plane, it can be shown that:

@@

The strain energy density function takes different forms for different materials. It is often expressed as a linar combination of unknown coefficients (determined experimentally) and principal invariants of Green strain tensor $\mathbf{E}$, left Cauchy-Green strain tensor $\mathbf{B}$ and deformation gradient tensor $\mathbf{F}$ (see section \ref{chap2:strainInvariants} for more information on invariants). The constitutive equations of  two well-known non-linear elastic materials (Mooney-Rivlin and neo-Hookean) will now be introduced.

@@

If continuum mechanics does teach us how to model idealised continuous structures, modelling soft tissues remains very complex. Indeed, in practice soft tissues are complex structures: they do not consist of a simple continuous piece of material, they are made of various constituents. Mechanical properties of soft tissues not only depends on the properties of these constituents, but how they are arranged relative to each other. This elaborate arrangement often leads to material properties for the global structure that are non-homogenous, non-linear, anisotropic and even viscoelastic.

>>>>

\section{Description of motion} 	\label{chap2:descriptionMotion}
Let us consider a body $\textbf{\textit{B}}$ of known geometry in a three-dimensional Euclidian space $\field{R}^3$. For a given geometry and loading, $\textbf{\textit{B}}$ will undergo a set of macroscopic changes which is called \emph{deformation}. The region of space occupied by the body at a given time $t$ is termed a \emph{configuration}. A change in the configuration of a continuum body results in a displacement. The displacement of a body has two components: a rigid-body displacement and a deformation. A rigid-body displacement consists of a simultaneous translation and rotation of the body without changing its shape or size. In contrast, a change in shape and\slash or size of the body from an initial configuration $\kappa_{0}$ to a new configuration $\kappa$ is called a deformation. This new configuration $\kappa$ may be designated by the \emph{current} or \emph{deformed configuration}. \OFF

@@

	\subsection{Deformation gradient tensor}
The displacement field tells us how a particle displaces from the reference to the deformed configuration. More interestingly, we would like to know how a material line would deform (stretch and rotation) within this displacement field. Since the length of a material line $d\mathbf{X}$ can change when going to the deformed configuration as well as its orientation, we can say that $d\mathbf{X}$ deforms into $d\mathbf{x}$. We now seek the relation existing between $d\mathbf{x}$ in the deformed configuration and $d\mathbf{X}$ of the reference configuration. \OFF

@@

\section{Constitutive equations}
So far, all the relations that we have established are valid for every continuum. Indeed, the derivations carried out so far made no mention of any material. Therefore, we know that these relations are not sufficient to describe the response of a material to a given loading. Moreover, because the stresses result from the deformation of the material, they may be expressed in terms of some measure of this deformation such as the strain. This relationship, which obviously depends on the type of material under consideration, is what we call a \emph{constitutive equation}. Constitutives equations are an attempt to mathematically characterise the behaviour of materials observed from experimental results. \OFF

@@

In addition, if there exists a \emph{strain energy density} function $U_0(\varepsilon)$ such that 
\begin{equation}
\label{chap2:energy}
\boldsymbol \sigma = \pd{U_0}{\varepsilon},
\end{equation}
the material is said to be hyperelastic. However, for \emph{incompressible} materials the stress tensor is not completely determined by the deformation. The hydrostatic pressure\footnote{pressure exerted by a fluid at equilibrium due to the force of gravity} affects the stress and the above equation is then written as:
\begin{equation}
\label{chap2:energy2}
\boldsymbol \sigma = -p \mathbf{I} + \pd{U_0}{\varepsilon},
\end{equation}
where $p$ is the hydrostatic pressure. In order to develop a mathematical model of an hyperelastic material, $U_0$ is expanded in Taylor's series about $\varepsilon = 0$:
\begin{equation}
U_0 = C_0 + C_{ij}\varepsilon_{ij} + \frac{1}{2!}\hat{C}_{ijkl}\varepsilon{ij}\varepsilon_{kl} + \frac{1}{3!}\hat{C}_{ijklmn}\varepsilon_{ij}\varepsilon_{kl}\varepsilon_{mn} + \ldots , 
\end{equation}
where $C_0$, $C_{ij}$ and $\hat{C}$ are material stiffnesses. The form of $U_0$ may varied for different materials. A material is said to be \emph{linear} if the relationship between strain and stress is also linear (see \fig{chap2:fig-linear}). Based on this definition, $U_0$ is a cubic or higher-order function of the strains for non-linear materials whereas $U_0$ is a quadratic function of strain for linear materials. We will start by discussing linear materials for the case of infinitesimal deformations (sections \ref{chap2:Hooke} to \ref{chap2:isotropic}). Consequently, we will not make any distinction between the various measures of stress and strain and use $\boldsymbol \sigma$ and $\boldsymbol \varepsilon$ for stress and strain tensors respectively. 
%
\begin{figure}[ht]
\centering
\includegraphics[width=7cm]{chapter2/linear.pdf}
\caption {The relationship between strain and stress is often non-linear (blue). However, it may be approximated as a linear relationship for small deformations (red). }
\label{chap2:fig-linear}
\end{figure}

@@

Most often, the material properties are determined in a laboratory in terms of the engineering constants such as Young's modulus, shear modulus, and so on. These constants are measured using simple tests like uni-axial tension test or pure shear test. Because of their direct and obvious physical meaning, engineering constants are used in place of the more abstract stiffness coefficients $C_{ijkl}$. Therefore, we will now reformulate \eqref{chap2:orthotropic} with respect to those engineering constants. Noting $E_i$ the Young's modulus in the direction $x_i$ and $G_{ij}$ the shear moduli in the $x_i-x_j$ plane, it can be shown that:

@@

The strain energy density function takes different forms for different materials. It is often expressed as a linear combination of unknown coefficients (determined experimentally) and principal invariants of Green strain tensor $\mathbf{E}$, left Cauchy-Green strain tensor $\mathbf{B}$ and deformation gradient tensor $\mathbf{F}$ (see section \ref{chap2:strainInvariants} for more information on invariants). The constitutive equations of  two well-known non-linear elastic materials (Mooney-Rivlin and neo-Hookean) will now be introduced.

@@

If continuum mechanics does teach us how to model idealised continuous structures, modelling soft tissues remains very complex. Indeed, in practice soft tissues are complex structures: they do not consist of a simple continuous piece of material, they are made of various constituents. Mechanical properties of soft tissues not only depends on the properties of these constituents, but how they are arranged relative to each other. This elaborate arrangement often leads to material properties for the global structure that are non-homogeneous, non-linear, anisotropic and even viscoelastic.

### 2010-10-22 Thesis/Chapter3.tex
- Update 22 October -
Almost done...



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@188 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

	\subsection{Meshing process}
The first step in the finite element method is to create a mesh of the domain to study. Mesh generation is a very important task and can be very time consuming. The domain has to be meshed properly into elements of specific shapes. All the elements together form the entire domain of the problem without any gap or overlapping. For example, triangles or quadrilaterals can be used in two dimensions, and tetrahedra and hexahedra in three dimensions. Information, such as \emph{element connectivity}, must be created during the meshing process for later use in the formulation of the FEM equations. The number of elements into which the domain is divided in a problem depends mainly on the geometry of the domain and on the desired accuracy of the solution. Usually, the number of elements increases with the complexity of the geometry. For instance, if one part of the domain is thiner (see \fig{chap3:fig-errorWithSubdivision} (c)), the size of the elements must be reduced in order to tile this particular part, which increases the total number of elements, and hence the number of algebraic equations to solve. However, adding elements is sometimes desirable. Indeed, increasing the number of elements tends to get an approximate solution closer to the exact analytical solution (as reducing the width of the rectangles allows for a better approximation of the area under the curve). Elements are classically added in the particular regions of interest for a given problem, if this information is known a priori. A trade-off between accuracy and computational time must be found. This is why the meshing process on the problem domain must be carried out carefully. One needs to create a mesh which gives an accurate enough solution for the desired application while restraining the computational time according to the time constraints of the problem. As an example, finding the maximum load that can sustain a bridge in structural mechanics demands a high degree of accuracy, no matter how much time it is needed to compute the solution. Conversely, organ deformation in medical training simulators must be computed at an interactive rate so that no apparent delay can be observed between the manipulation of a given organ and its deformation. It does not mean that precision is not required, simply that the time constraints will necessarily limit the accuracy of the solution.

@@

\noindent \textbf{Remark.}
It is worth noting that \eqref{chap3:shapeFunctions} can be easily extented to three-dimensional problems. Let us consider a Cartesian coordinate system and we denote the position of node $ i $ $ \mathbf{x}_i = [x_{i1}, x_{i2}, x_{i3}]^T$. The position of a point $ \mathbf{x} = [x_{1}, x_{2}, x_{3}]^T$ within the element may then be interpolated by applying \eqref{chap3:shapeFunctions} to each component of the position:
\begin{align}
x_1 &= \sum_{i=1}^m x_{i1} \hat{\psi}_i^e(\xi) \notag \\
x_2 &= \sum_{i=1}^m x_{i2} \hat{\psi}_i^e(\xi) \\
x_3 &= \sum_{i=1}^m x_{i3} \hat{\psi}_i^e(\xi) \notag
\end{align}
If we construct the vector $ \hat{\mathbf{x}}^e $ of all nodal positions $ \mathbf{x}_i $ of the element, we can define a shape function matrix $ \mathbf{H} $ such as:
\begin{equation}
\label{chap3:H}
\mathbf{x} = \mathbf{H} \hat{\mathbf{x}}^e
\end{equation}
where $ \mathbf{H} $ is a $ 3 \times m $ matrix built from submatrices $ \mathbf{H}_i $
\begin{equation}
\mathbf{H}_i = 
	\begin{bmatrix}
		\hat{\psi}_i^e & 0 & 0 \\
		0 & \hat{\psi}_i^e & 0 \\
		0 & 0 & \hat{\psi}_i^e
	\end{bmatrix}
\quad i = 1, 2, \ldots, m.
\end{equation}

	\subsection{A particular case: isoparametric elements}
% \noindent \textbf{Some remarks.}
Most of the time, we choose to interpolate the solution and the element geometry with the same interpolation functions. In this case, the elements are said to be \emph{isoparametric}. Under this configuration, natural coordinates appear as parameters that define the shape functions. Note that the shape functions describe the geometry and the solution with the same degree of interpolation. Because isoparametric elements are very common in practice, the use of the phrase \emph{shape functions} is often extented to denote the interpolation functions employed to approximate the solution.

@@

	\subsection{Time dependence}
In all the equations of the finite element method that we derived so far, we have not taken time into account. For some problem, this is not an issue, the motion is sufficiently slow to assume that the dynamic effects (such as damping effect) are neglectable. However, this assumption is not always valid, particulary in our case where we want to model organ deformation. The deformation obviously depends on time. Finite element models of time-dependent problems can be developed in two alternatives ways \citep{Reddy93}:
\begin{description}
\item[(a)] coupled formulation in which the time $ t $ is an additional coordinate along with the spatial coordinate $ x $
\item[(b)] decoupled formulation where time and spatial variations are assumed to be separable.
\end{description}
Thus, the approximation of the solution $ u $ takes one of these two forms:
\begin{align}
u(x, t) \approx u^e(x, t) &= \sum_{i=1}^n \hat{u}^e_i \hat{\psi}^e_i(x, t)  \quad \text{ (coupled formulation)} \label{chap3:approxTime1}\\
u(x, t) \approx u^e(x, t) &= \sum_{i=1}^n u^e_i(t) \psi^e_i(x) \quad \text{(decoupled formulation)} \label{chap3:approxTime2}
\end{align}
where $ \hat{\psi}^e_i(x, t) $ are time-space interpolation functions, $ \hat{u}^e_i $ are the nodal values that are independent of $ x $ and $ t $, $ \psi^e_i(x) $ are the usual interpolation functions in spatial coordinate $ x $ only and the nodal values $ u^e_i(t) $ are functions of time $ t $ only. Coupled finite element formulations are not common and they have not been adequately studied. Hence, we consider the decoupled formulation only. Of course, the assumption that the time and spatial variations are separable is not valid in general. However, with sufficiently small time steps, it is possible to obtain accurate solutions nevertheless.

@@

\section{Solution of global problem}
The complete and global system of equations has now been established. In this section, we will examine the techniques commonly used for solving these equations. In the general case of a dynamic system of equations, the equations are not yet algebraic and still depends on derivatives of the unknown displacements with respect to time. Therefore, an additional step needs to be performed before using a solver by integrating the equations over time. As we will see shortly, the different methods of time integration are classified into two main schemes: \emph{explicit} or \emph{implicit} integration. We begin by introducing the simplest method, the Euler method to explain the concept of an explicit method. Then the central difference method will be detailled. We then describe a few implicit methods, allowing us to highlight the relative advantages of the two types of methods. Finally, the static case will be solved by emphasizing the similarities with implicit systems through the critical step of linearisation of the governing equations \citep{Belytschko00}.

@@

\ON
\paragraph*{Preconditioners}
Unfortunately, it is common for the matrix $ \mathbf{A} $ not to be well conditioned. Consequently, the convergence of the chosen iterative method is often slow. In order to achieve an accelerated convergence, we may use a \emph{preconditioner}. Although the formation and the study of preconditioners are out of the scope of this PhD thesis, preconditioners are a key element for solving systems of equations. To be effective, a preconditioner $ \mathbf{P} $ should be a good approximation of  $ \mathbf{A} $ and the following preconditioned system must be easier to solve than the original system \eqref{chap3:linearEquation}:
\begin{equation}
\label{chap3:preconditionedSystem}
\mathbf{P}^{-1} \mathbf{A} \mathbf{x} = \mathbf{P}^{-1} \mathbf{b}.
\end{equation}
Two criteria should also be considered when forming a preconditioner: (1) the ease with which it may be evaluated and (2) its memory requirements. The easiest preconditioner would be the identity since $ \mathbf{P}^{-1} = \mathbf{I} $ in this case. Of course, such a conditioner leads to the resolution of the original system and is therefore totally useless. In constrast, the choice of $ \mathbf{P} = \mathbf{A} $ leads to the solution of the system \eqref{chap3:linearEquation} in one single iteration. In practice, an optimal preconditioner is to be found between these two extremes to minimise the number of iterations required by the solver while keeping the evaluation of the preconditioner simple. Finding a good preconditioner for all types of problems is difficult. A preconditioner is only useful if the overall time spent on both the formation of  $ \mathbf{P} $ and the resolution of \eqref{chap3:preconditionedSystem} is less than the time spent at solving \eqref{chap3:linearEquation} alone.

One of the simplest preconditioner is the \emph{Jacobi preconditioner} where the preconditioner is chosen to be the diagonal of the matrix $ \mathbf{A} $. This choice is efficient in cases where $ \mathbf{A} $ is diagonally dominant. Sparse approximations of LU or Cholesky decompositions may also be used as preconditioners (see sections \ref{chap3:LUdecompo} and \ref{chap3:CholeskyDecompo} for more details). 
\OFF

>>>>

	\subsection{Meshing process}
The first step in the finite element method is to create a mesh of the domain to study. Mesh generation is a very important task and can be very time consuming. The domain has to be meshed properly into elements of specific shapes. All the elements together form the entire domain of the problem without any gap or overlapping. For example, triangles or quadrilaterals can be used in two dimensions, and tetrahedra and hexahedra in three dimensions. Information, such as \emph{element connectivity}, must be created during the meshing process for later use in the formulation of the FEM equations. The number of elements into which the domain is divided in a problem depends mainly on the geometry of the domain and on the desired accuracy of the solution. Usually, the number of elements increases with the complexity of the geometry. For instance, if one part of the domain is thinner (see \fig{chap3:fig-errorWithSubdivision} (c)), the size of the elements must be reduced in order to tile this particular part, which increases the total number of elements, and hence the number of algebraic equations to solve. However, adding elements is sometimes desirable. Indeed, increasing the number of elements tends to get an approximate solution closer to the exact analytical solution (as reducing the width of the rectangles allows for a better approximation of the area under the curve). Elements are classically added in the particular regions of interest for a given problem, if this information is known a priori. A trade-off between accuracy and computational time must be found. This is why the meshing process on the problem domain must be carried out carefully. One needs to create a mesh which gives an accurate enough solution for the desired application while restraining the computational time according to the time constraints of the problem. As an example, finding the maximum load that can sustain a bridge in structural mechanics demands a high degree of accuracy, no matter how much time it is needed to compute the solution. Conversely, organ deformation in medical training simulators must be computed at an interactive rate so that no apparent delay can be observed between the manipulation of a given organ and its deformation. It does not mean that precision is not required, simply that the time constraints will necessarily limit the accuracy of the solution.

@@

\noindent \textbf{Remark.}
It is worth noting that \eqref{chap3:shapeFunctions} can be easily extended to three-dimensional problems. Let us consider a Cartesian coordinate system and we denote the position of node $ i $ $ \mathbf{x}_i = [x_{i1}, x_{i2}, x_{i3}]^T$. The position of a point $ \mathbf{x} = [x_{1}, x_{2}, x_{3}]^T$ within the element may then be interpolated by applying \eqref{chap3:shapeFunctions} to each component of the position:
\begin{align}
x_1 &= \sum_{i=1}^m x_{i1} \hat{\psi}_i^e(\xi) \notag \\
x_2 &= \sum_{i=1}^m x_{i2} \hat{\psi}_i^e(\xi) \\
x_3 &= \sum_{i=1}^m x_{i3} \hat{\psi}_i^e(\xi) \notag
\end{align}
If we construct the vector $ \hat{\mathbf{x}}^e $ of all nodal positions $ \mathbf{x}_i $ of the element, we can define a shape function matrix $ \mathbf{H} $ such as:
\begin{equation}
\label{chap3:H}
\mathbf{x} = \mathbf{H} \hat{\mathbf{x}}^e
\end{equation}
where $ \mathbf{H} $ is a $ 3 \times m $ matrix built from submatrices $ \mathbf{H}_i $
\begin{equation}
\mathbf{H}_i = 
	\begin{bmatrix}
		\hat{\psi}_i^e & 0 & 0 \\
		0 & \hat{\psi}_i^e & 0 \\
		0 & 0 & \hat{\psi}_i^e
	\end{bmatrix}
\quad i = 1, 2, \ldots, m.
\end{equation}

	\subsection{A particular case: isoparametric elements}
% \noindent \textbf{Some remarks.}
Most of the time, we choose to interpolate the solution and the element geometry with the same interpolation functions. In this case, the elements are said to be \emph{isoparametric}. Under this configuration, natural coordinates appear as parameters that define the shape functions. Note that the shape functions describe the geometry and the solution with the same degree of interpolation. Because isoparametric elements are very common in practice, the use of the phrase \emph{shape functions} is often extended to denote the interpolation functions employed to approximate the solution.

@@

	\subsection{Time dependence}
In all the equations of the finite element method that we derived so far, we have not taken time into account. For some problem, this is not an issue, the motion is sufficiently slow to assume that the dynamic effects (such as damping effect) are neglectable. However, this assumption is not always valid, particularly in our case where we want to model organ deformation. The deformation obviously depends on time. Finite element models of time-dependent problems can be developed in two alternatives ways \citep{Reddy93}:
\begin{description}
\item[(a)] coupled formulation in which the time $ t $ is an additional coordinate along with the spatial coordinate $ x $
\item[(b)] decoupled formulation where time and spatial variations are assumed to be separable.
\end{description}
Thus, the approximation of the solution $ u $ takes one of these two forms:
\begin{align}
u(x, t) \approx u^e(x, t) &= \sum_{i=1}^n \hat{u}^e_i \hat{\psi}^e_i(x, t)  \quad \text{ (coupled formulation)} \label{chap3:approxTime1}\\
u(x, t) \approx u^e(x, t) &= \sum_{i=1}^n u^e_i(t) \psi^e_i(x) \quad \text{(decoupled formulation)} \label{chap3:approxTime2}
\end{align}
where $ \hat{\psi}^e_i(x, t) $ are time-space interpolation functions, $ \hat{u}^e_i $ are the nodal values that are independent of $ x $ and $ t $, $ \psi^e_i(x) $ are the usual interpolation functions in spatial coordinate $ x $ only and the nodal values $ u^e_i(t) $ are functions of time $ t $ only. Coupled finite element formulations are not common and they have not been adequately studied. Hence, we consider the decoupled formulation only. Of course, the assumption that the time and spatial variations are separable is not valid in general. However, with sufficiently small time steps, it is possible to obtain accurate solutions nevertheless.

@@

\section{Solution of global problem}
The complete and global system of equations has now been established. In this section, we will examine the techniques commonly used for solving these equations. In the general case of a dynamic system of equations, the equations are not yet algebraic and still depends on derivatives of the unknown displacements with respect to time. Therefore, an additional step needs to be performed before using a solver by integrating the equations over time. As we will see shortly, the different methods of time integration are classified into two main schemes: \emph{explicit} or \emph{implicit} integration. We begin by introducing the simplest method, the Euler method to explain the concept of an explicit method. Then the central difference method will be detailed. We then describe a few implicit methods, allowing us to highlight the relative advantages of the two types of methods. Finally, the static case will be solved by emphasizing the similarities with implicit systems through the critical step of linearisation of the governing equations \citep{Belytschko00}.

@@

\ON
\paragraph*{Pre-conditioners}
Unfortunately, it is common for the matrix $ \mathbf{A} $ not to be well conditioned. Consequently, the convergence of the chosen iterative method is often slow. In order to achieve an accelerated convergence, we may use a \emph{pre conditioner}. Although the formation and the study of pre-conditioners are out of the scope of this PhD thesis, pre-conditioners are a key element for solving systems of equations. To be effective, a preconditioner $ \mathbf{P} $ should be a good approximation of  $ \mathbf{A} $ and the following preconditioned system must be easier to solve than the original system \eqref{chap3:linearEquation}:
\begin{equation}
\label{chap3:preconditionedSystem}
\mathbf{P}^{-1} \mathbf{A} \mathbf{x} = \mathbf{P}^{-1} \mathbf{b}.
\end{equation}
Two criteria should also be considered when forming a preconditioner: (1) the ease with which it may be evaluated and (2) its memory requirements. The easiest pre-conditioner would be the identity since $ \mathbf{P}^{-1} = \mathbf{I} $ in this case. Of course, such a conditioner leads to the resolution of the original system and is therefore totally useless. In constrast, the choice of $ \mathbf{P} = \mathbf{A} $ leads to the solution of the system \eqref{chap3:linearEquation} in one single iteration. In practice, an optimal pre-conditioner is to be found between these two extremes to minimise the number of iterations required by the solver while keeping the evaluation of the pre-conditioner simple. Finding a good pre-conditioner for all types of problems is difficult. A pre-conditioner is only useful if the overall time spent on both the formation of  $ \mathbf{P} $ and the resolution of \eqref{chap3:preconditionedSystem} is less than the time spent at solving \eqref{chap3:linearEquation} alone.

One of the simplest pre-conditioner is the \emph{Jacobi pre-conditioner} where the pre-conditioner is chosen to be the diagonal of the matrix $ \mathbf{A} $. This choice is efficient in cases where $ \mathbf{A} $ is diagonally dominant. Sparse approximations of LU or Cholesky decompositions may also be used as pre-conditioners (see sections \ref{chap3:LUdecompo} and \ref{chap3:CholeskyDecompo} for more details). 
\OFF
